Stressing Bank Profitability for Interest Rate Risk

### 2013-07-08 LATEX/slides/nim.slides.tex
updated slides

\begin{frame}
\frametitle{Net interest margins (for the top 25 bank holding companies, assessed quarterly)}

@@

\begin{frame}
\frametitle{Why study NIMs} 
\begin{itemize}
\item \vspace{0.0in} Losses from banks' earning depressed or negative NII is an important source of risk to banks and the financial sector.
\item \vspace{0.1in} The U.S. savings and loans crisis was associated with NII and NIMs turning negative in the sector.
\end{itemize}
\begin{center}
\includegraphics[height=4.4cm,width=5.9cm]{NIMs_Thrifts_CROP.pdf}
\end{center}
\vspace{-0.2in} \hspace{1.0in} {\tiny{Source: FR Bulletin, February 1996.}}\normalsize
\end{frame}

\begin{frame}
\frametitle{Why study NIMs, continued} 
\begin{itemize}
\item \vspace{0.0in} A few other banking crises have also been associated with depressed or negative NIMs; including,       
\begin{itemize}
\item \vspace{0.125in} The Nordic banking crises (in Finland, Sweden, and Norway) of the late 1980s/early 1990s; and, 
\item \vspace{0.125in} The U.K. secondary banking of 1973-75. 
\end{itemize}
\item \vspace{0.25in} NII is an important part of bank profitability.
\begin{itemize}
\item \vspace{0.125in} Bank equity analysis direct considerable attention to NII and NIMs in evaluating future bank profitability.
\item \vspace{0.125in} This reflects the fact that NII and NIMs are less volatile than other components of bank revenue.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Key variables for modeling NIMs}
\begin{itemize}
\item \vspace{0.0in} The slope of the Treasury yield curve. 
\begin{itemize}
\item \vspace{0.05in} Reflects banks' return on maturity-transformation serivces -- one of the key services provided by banks.  
\end{itemize}
\item \vspace{0.1in} The level of short-term interest rates. 
\begin{itemize}
\item \vspace{0.05in} Indirectly reflects banks' return on transactions services -- another key service provided by banks.
\item \vspace{0.05in} The level of the short rate puts an upper limit on how much banks can earn from transactions services.  
\end{itemize}
\item \vspace{0.1in} The use of these variables is common in the macro-banking literature modeling NIMs.
\begin{itemize}
\item \vspace{0.05in} Kovner and Vickery (2012)
\item \vspace{0.05in} Covas, Rump, and Zakrajsek (2012)
\item \vspace{0.05in} English (2002)
\item \vspace{0.05in} English, Van den Heuvel, and Zakrajsek (2012)
\end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{0.1in}\hspace{0.1in}
\includegraphics[height=3.6cm,width=4.8cm]{nims_slope_crop.pdf}  \hspace{0.25in}
\includegraphics[height=3.6cm,width=4.8cm]{nims_level_crop.pdf}
\begin{itemize}
\item\vspace{0.1in} NIMs increase when the yield-curve steepens.
\begin{itemize}
\item \vspace{0.025in} Reflects the increased return to maturity transformation.
\end{itemize}
\item \vspace{0.075in} Changes in the short rate drive changes in the slope of the yield curve. 
\item \vspace{0.075in} There is an evident downward trend in the time series. 
\begin{itemize}
\item \vspace{0.025in} This may be tied to the lower level of interest rates.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Net interest margins:  A few quirks in the data}

@@

ADD SOMETHING ON WHAT IS BEING CAPTURED AS A SHADOW BANK

@@

>>>>

\begin{frame}
\frametitle{NIMs (for the top 25BHCs, assessed quarterly)}

@@

\begin{frame}
\frametitle{Why study NIMs}
\begin{itemize}
\item \vspace{0.0in} Losses from banks' earnings depressing NII is an important source of risk to banks and the financial sector.
\item \vspace{0.1in} The U.S. savings and loans crisis was associated with NII and NIMs turning negative in the thrift sector.
\end{itemize}
\begin{center}
\includegraphics[height=4.4cm,width=5.9cm]{NIMs_Thrifts_CROP.pdf}
\end{center}
\vspace{-0.2in} \hspace{1.0in} {\tiny{Source: FR Bulletin, February 1996.}}\normalsize
\end{frame}

\begin{frame}
\frametitle{Why study NIMs, continued}
\begin{itemize}
\item \vspace{0.0in} A few other banking crises have also been associated with depressed or negative NIMs; including,
\begin{itemize}
\item \vspace{0.125in} The Nordic banking crises (in Finland, Sweden, and Norway) of the late 1980s/early 1990s; and,
\item \vspace{0.125in} The U.K. secondary banking of 1973-75.
\end{itemize}
\item \vspace{0.25in} NII is an important part of bank profitability.
\begin{itemize}
\item \vspace{0.125in} Bank equity analysts direct considerable attention to NII and NIMs in evaluating future bank profitability.
\item \vspace{0.125in} This reflects the fact that NII and NIMs are less volatile than other components of bank revenue.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Key variables for modeling NIMs}
\begin{itemize}
\item \vspace{0.0in} The slope of the Treasury yield curve.
\begin{itemize}
\item \vspace{0.05in} Reflects banks' return on maturity-transformation serivces -- one of the key services provided by banks.
\end{itemize}
\item \vspace{0.1in} The level of short-term interest rates.
\begin{itemize}
\item \vspace{0.05in} Indirectly reflects banks' return on transactions services -- another key service provided by banks.
\item \vspace{0.05in} The level of the short rate puts an upper limit on how much banks can earn from transactions services.
\end{itemize}
\item \vspace{0.1in} The 10-year yield less 3-month rate and the 3-month rate are commonly used in the macro-banking NIM literature.
\begin{itemize}
\item \vspace{0.05in} Kovner and Vickery (2012)
\item \vspace{0.05in} Covas, Rump, and Zakrajsek (2012)
\item \vspace{0.05in} English (2002)
\item \vspace{0.05in} English, Van den Heuvel, and Zakrajsek (2012)
\end{itemize}
\end{itemize}
\end{frame}

@@

\vspace{0.1in}\hspace{0.1in}
\includegraphics[height=3.6cm,width=4.8cm]{nims_slope_crop.pdf}  \hspace{0.25in}
\includegraphics[height=3.6cm,width=4.8cm]{nims_level_crop.pdf}
\begin{itemize}
\item\vspace{0.1in} NIMs increase when the yield-curve steepens.
\begin{itemize}
\item \vspace{0.025in} Reflects the increased return to maturity transformation.
\end{itemize}
\item \vspace{0.075in} Changes in the short rate drive changes in the slope of the yield curve.
\item \vspace{0.075in} There is an evident downward trend in the NIM time series.
\begin{itemize}
\item \vspace{0.025in} This may be tied to the lower level of interest rates.
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{NIMs:  A few quirks in the data}

@@

\begin{itemize}
\item \vspace{0.0in} The most relevant source of competition compressing NIMs seems to have come from the shadow banking sector, which,
\begin{itemize}
\item \vspace{0.1in} Offers alternatives to deposits; \emph{e.g.}, MMMFs.
\item \vspace{0.1in} Offers alternatives to bank loans, \emph{e.g.}, from finance companies.
\end{itemize}
\item \vspace{0.125in} We measure competition by the share of intermediated credit accounted for by shadow banking sector.
\item \vspace{0.125in} Following Adrian, Monch, and Shin (2010) we define:
\begin{itemize}
\item \vspace{0.1in} The shadow banking sector as funding corporations, mortgage pools, finance companies, ABS issuers, and broker-dealers; and,
\item \vspace{0.1in} The traditional banking sector as commercial banks, saving institutions, and credit unions.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Competition and NIMs}

\hspace{0.65in}\includegraphics[height=5.4cm,width=7.2cm]{nims_competition_crop.pdf}

@@

\begin{frame}
\frametitle{CCAR-2013 Scenarios}
\vspace{-0.4in}
\hspace{0.65in}\includegraphics[height=10.4cm,width=8cm]{figure_ccar_2013.pdf}
